\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{colorlinks=true, urlcolor=blue, linkcolor=black}

\title{Competitive Districts (Simplified Applet): Technical Reference}
\author{}
\date{}

\newcommand{\op}[1]{\operatorname{#1}}
\newcommand{\eps}{\varepsilon}
\newcommand{\R}{\mathbb{R}}

\begin{document}
\maketitle

This document is the source of truth for the model implemented in
\texttt{index.html} (the default, simplified applet; the fuller exploratory
version with every raw parameter exposed lives in \texttt{full.html}).
The applet exposes seven plain-English knobs to the user; each maps onto a
subset of the underlying parameters described below.
All ideology / partisanship quantities are in \emph{percentage points}.

\section{Model setup}

The chamber consists of $2m+1$ districts indexed by
$i \in \{0, 1, \ldots, 2m\}$, ordered from most Democratic to most Republican
by their relative partisanship $d_i \in [-100, 100]$.
$d_i = 0$ corresponds to a tied (50--50) district.

\subsection{District partisanship distribution}

Districts are placed \emph{deterministically} at the quantiles of a Gaussian
$\mathcal{N}(\mu_{\text{district}}, \sigma_{\text{district}}^2)$ truncated to
$(0, 100]$, with the negative half being the mirror image. Concretely, for
$i = 0, \ldots, m-1$:
\[
  q_i = \frac{i + \tfrac{1}{2}}{m},
  \qquad
  u_i = \Phi\!\left(\frac{-\mu_{\text{district}}}{\sigma_{\text{district}}}\right)
        + q_i \left[
          \Phi\!\left(\frac{100-\mu_{\text{district}}}{\sigma_{\text{district}}}\right)
        - \Phi\!\left(\frac{-\mu_{\text{district}}}{\sigma_{\text{district}}}\right)
        \right],
\]
\[
  d_{m+1+i} \;=\; \mu_{\text{district}} + \sigma_{\text{district}} \, \Phi^{-1}(u_i),
  \qquad
  d_{m-1-i} \;=\; -d_{m+1+i},
  \qquad d_m = 0.
\]
Here $\Phi$ is the standard normal CDF. The result is a smooth, symmetric,
bimodal partisanship distribution, identical across simulation runs.

\subsection{Candidate ideologies}

For each district $i$ the two candidates' ideologies are sampled
independently:
\begin{align}
  c^D_i &= \beta^D_{\text{swing}} \log\!\frac{100}{|d_i|}
         + \beta^D_{\text{compet}} \log\!\frac{100}{|d_i + v|}
         + \eps^D_i,
         &\eps^D_i &\sim \mathcal{N}(\mu^D, \sigma_D^2),
  \\
  c^R_i &= -\beta^R_{\text{swing}} \log\!\frac{100}{|d_i|}
         - \beta^R_{\text{compet}} \log\!\frac{100}{|d_i + v|}
         + \eps^R_i,
         &\eps^R_i &\sim \mathcal{N}(\mu^R, \sigma_R^2).
\end{align}
$|d_i|$ is floored at $0.1$ before the logarithm to avoid blow-up at the median
($d_m = 0$) district. Note that $\log(100/|d|)$ is large ($\approx 6.9$) for
very competitive districts and $\approx 0$ for very safe ones, so positive
$\beta$ pulls each party's candidate \emph{toward zero} (more moderate) in
swingier districts. $v$ is the popular vote shift in percentage points (R--D).

\subsection{Election outcome}

Each district's winner is a Bernoulli draw with
\[
  P(\text{R wins district } i)
  \;=\; \sigma\!\left(
    \frac{v + d_i - w_{\text{mod}}\,(c^D_i + c^R_i)}{\sigma_{\text{noise}}}
  \right),
  \qquad
  \sigma(x) = \frac{1}{1 + e^{-x}}.
\]
The realized rep ideology is $r_i = c^R_i$ if R wins, $r_i = c^D_i$ otherwise.
$w_{\text{mod}}$ controls how strongly voters punish the more-extreme
\emph{average} of the two candidates relative to district lean.

\subsection{Median representative}

After every district is decided, sort the elected reps by ideology and take
the rank-$m$ value:
\[
  \tilde{r} \;=\; \op{sort}_{\text{asc}}(r_0, \ldots, r_{2m})_{[m]},
  \qquad
  P(\tilde{r}) \in \{D, R\}.
\]
The applet runs $N_{\text{sim}}$ such simulations and visualizes the
empirical distributions of $\tilde{r}$ and of the per-district outcomes.

\section{Fixed constants}

The following are hard-coded in the simplified applet (not user-adjustable):

\begin{center}
\begin{tabular}{ll}
\toprule
$m$ & $217$ \quad (so $2m+1 = 435$ districts) \\
$N_{\text{sim}}$ & $1000$ \\
$\mu_{\text{district}}$ & $20$ \\
$\sigma_{\text{noise}}$ & $10$ \\
\bottomrule
\end{tabular}
\end{center}

\section{User-facing slider $\to$ parameter map}

Each slider's numeric value $s$ feeds into one or more underlying parameters.

\subsection*{Popular vote \% (R--D)}
Direct map: $v \leftarrow s$, with $s \in [-20, 20]$, default $0$.

\subsection*{District competitiveness}
$\sigma_{\text{district}} \leftarrow s$, with $s \in [1, 15]$, default $10$.
Smaller values concentrate districts near the safe-D and safe-R modes; larger
values spread mass into the swing range.

\subsection*{Democratic candidate ambient moderation}
A single slider drives both $\sigma_D$ and $\mu^D$ via a linear coupling
chosen so that ``more extreme on average $\Leftrightarrow$ lower variance'':
\[
  \sigma_D \;\leftarrow\; s,
  \qquad
  \mu^D \;\leftarrow\; -120 + 2 s,
  \qquad s \in [5, 15], \text{ default } 10.
\]
At $s=5$: $(\mu^D, \sigma_D) = (-110, 5)$ -- D's are extreme on average and
tightly clustered. At $s=15$: $(-90, 15)$ -- D's are slightly more moderate
on average but more dispersed.

\subsection*{Republican candidate ambient moderation}
Mirror-symmetric:
\[
  \sigma_R \;\leftarrow\; s,
  \qquad
  \mu^R \;\leftarrow\; +120 - 2 s,
  \qquad s \in [5, 15], \text{ default } 10.
\]

\subsection*{Democratic candidate intentional moderation}
Single slider drives both $\beta^D$ coefficients at a $1{:}1$ ratio:
\[
  \beta^D_{\text{swing}} = \beta^D_{\text{compet}} \;\leftarrow\; s,
  \qquad s \in [-5, 5], \text{ default } 1.
\]
Positive values make Democratic candidates moderate (drift toward $0$) in
competitive districts.

\subsection*{Republican candidate intentional moderation}
\[
  \beta^R_{\text{swing}} = \beta^R_{\text{compet}} \;\leftarrow\; s,
  \qquad s \in [-5, 5], \text{ default } 1.
\]

\subsection*{Candidate quality importance}
Direct map: $w_{\text{mod}} \leftarrow s$, with $s \in [-1, 2]$, default
$0.5$. Larger $w_{\text{mod}}$ means voters punish the more-extreme average
candidate more heavily relative to district lean.

\section{Symmetry}

At $v = 0$ with $\mu^R = -\mu^D$, $\sigma_R = \sigma_D$, $\beta^R = \beta^D$,
the model is exactly symmetric under the joint relabeling
$(D \leftrightarrow R,\; d \to -d,\; r \to -r)$, so the empirical distribution
of $\tilde{r}$ is symmetric around $0$ and $\Pr[P(\tilde{r}) = R] = \tfrac{1}{2}$.

\section{Change log}

\begin{itemize}
  \item \textbf{Initial simplified model.} Seven user-facing knobs; numeric
        readouts hidden for all sliders except popular vote; district
        partisanships placed deterministically at truncated-Gaussian quantiles;
        ambient-moderation sliders use the gentle linear coupling
        $\mu^{D/R} = \mp(120 - 2 s)$ on $s \in [5, 15]$.
  \item \textbf{Symbol-free chart and stats labels.} The applet UI no longer
        references $\tilde r$ or $d$ directly: chart titles, axis labels,
        hover templates, and stats card all use plain English (``ideology'',
        ``district partisanship'', ``median representative''). The percent
        sign is kept on district-partisanship axes (it is a meaningful unit:
        Republican vote share minus Democratic vote share, in percentage
        points), and is dropped from ideology axes (the ideology scale has no
        natural percent interpretation to a non-modeler). Internally the
        simulator continues to operate on the same dimensionless quantities
        described in the body of this document.
  \item \textbf{Chart layout polish.} Chart panels grew from 300 to 360 px
        tall; bottom margins enlarged and legends pushed below the plot area
        to prevent overlap with x-axis titles; the scatter chart's y-axis
        title was shortened to ``Rep ideology'' (without a directional
        suffix) since rotated text could not fit the long phrase at the
        chart height.
  \item \textbf{$N_{\text{sim}}$ raised from 400 to 1000.} Both the fixed
        constant and every prose mention in the applet (subtitle, footnote)
        were updated.
  \item \textbf{File rename for static hosting.} The simplified applet was
        renamed \texttt{simple.html $\to$ index.html} so that Fly.io,
        Netlify, GitHub Pages, etc.\ serve it as the default. The fuller
        exploratory applet was renamed \texttt{index.html $\to$ full.html}.
        No model behavior changed.
  \item \textbf{Popular vote slider redesign.} Range narrowed from
        $[-20, 20]$ to $[-15, 15]$. The numeric readout was replaced by a
        styled label that reads \texttt{R+$x$\%} when $v > 0$,
        \texttt{D+$x$\%} when $v < 0$, and \texttt{Tied} at $v = 0$, with
        the label colored red / blue / accent respectively. The editable
        number input was removed, so the slider now runs the full row width
        like the other six sliders.
  \item \textbf{Hint copy fix.} Both ambient-moderation hints changed
        ``but they vary more'' to ``and they vary more'' (the variance
        increase is concurrent with, not opposed to, the moderation shift).
  \item \textbf{Hint copy: drop ``slightly''.} Both ambient-moderation hints
        now read ``tend to be more moderate on average'' rather than
        ``tend to be slightly more moderate on average.''
\end{itemize}

\end{document}
